---
title: "Exploratory Analysis"
output: html_document
---

## Project title

Give your project a concise, interesting title that summarizes the entirety of your project

### Authors

Robert Bonglamphone, 

### Date

Winter 2022

### Abstract

No more than three sentences that summarize your project. Focus on the very most important aspects. For example: (1) "Our main question is .... This question is important because .... To address the question, we will ...." (2) "We are concerned with ..., because .... To address this concern, we plan to ...." (3) "Consider that .... This is important because .... Accordingly, we plan to ...."

GitHub site: https://info-201a-wi23.github.io/exploratory-analysis-bonglrob/

### Keywords

3-5 keywords that summarize your project.
(e.g., "Keywords: human physiology; bicycle exercise; Gen Z; times-series data")

### Introduction  

Briefly introduce your project.  Include 3-5 research questions. What motivates the questions? Why are they important? (at least 200 words)

### Related Work  

Describe your topic and related work in this space. You must include 3 citations to related work (URLs to similar work, high quality articles from the popular press, research papers, etc. ) Please use a standard citation style of your choice. (at least 200 words)

### The Dataset

Where did you find the data? Please include a link to the data source  
Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)

### Implications

Assuming you answer your research questions, briefly describe the expected or possible implications for technologists, designers, and policymakers. (at least 150 words)

### Limitations & Challenges
What challenges or limitations might you need to address with your project idea more broadly? Briefly discuss. (at least 150 words)

### Summary Information

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script

### Table

Include a table of aggregate information

Describe why you included the table and what information it reveals

### Chart 1

Describe why you chose this chart and what information it reveals

There are many causes of death in the United States, but my focus was to see what was
the main cause of death. Specifically, the audience should be able to draw a quick comparison between states and how they differ. Therefore, a map of the United States was the perfect fit. Also, there was only one of 2 causes of death that a state can be classified under, so a state would be the color of the leading cause of death that it represents. 

The map shows the top cause of death in the United States per state in 2017. Either a state's population main cause of death would be between heart disease and cancer. Overwhelmingly, heart disease is the main cause of the death between all the states. Roughly 1/4 of the states' main cause of death was cancer. 


```{r, echo = FALSE, code = readLines("chart1.R"), message = FALSE, warning = FALSE}
```

### Chart 2

Include a chart

Describe why you chose this chart and what information it reveals

Here's an example of how to run an R script inside an RMarkdown file:

```{r, echo = FALSE, code = readLines("chart1_example.R"), message = FALSE, warning = FALSE}
```


### Chart 3

Include a chart

Describe why you chose this chart and what information it reveals
